---
title: "SNOTEL clim aggregation and comparison with PRISM"
author: "Kelly Carscadden"
date: '2020-10-21'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(data.table)
library(naniar) #NA replacement
library(patchwork) #panel figures
theme_set(theme_classic())
```

Aims: Validate my use of PRISM data to predict flowering phenology at RMBL by:
1. Aggregating SNOTEL climate data that surrounds the study site
2. Comparing SNOTEL estimates to PRISM estimates (which are coarser in scale but have more years of data available)


Using: SNOTEL climate data from two sites that bracket RMBL (Mt Crested Butte, Schofield Pass), from 1986-2019

SNOTEL provides daily data for each year -> aggregate across sites and dates to get avg May-Jul ('mjj') temperature. Note, I also explored precipitation, but it was not recorded in the same way as PRISM, so wasn't comparable in the end.


```{r}
#read may, june, jul daily data from 2 SNOTEL stations
snotel <- 
    list.files(pattern = "*.csv") %>% 
    map_df(~fread(.)) #this cuts the 1st 4 rows (blank & metadata), and made the headers the headers

#remove unnecessary variables
snotel <- snotel %>% select(-c(Time, V10, `SNWD.I-1 (in)`, V11))

#fix colnames
colnames(snotel) <- c("site", "date", "wteq_i_in", "ppt_i_in",
          "tobs_i_C", "tmax_d_C", "tmin_d_C", "tavg_d_C")

#change missing data value from -99.9 to NA, before averaging
snotel <- snotel %>% 
  replace_with_na_all(condition = ~.x == -99.9) #slow step here

#spot check
summary(snotel$tmax_d_C) #lots of NAs, but now min/max seem plausible
str(snotel)

#split yyyy/mm/dd and convert inches -> cm for consistency with PRISM
snotel <- snotel %>% 
  separate(date, into=c("year", "mm", "dd"), sep = "-") %>%
  mutate(wteq_i_cm = wteq_i_in*2.54,
         ppt_i_cm = ppt_i_in*2.54,
         mm = as.numeric(mm),
         year = as.numeric(year)) #temps are already C
#make month, year numeric so easier to subset

#looking at raw data to check for outliers
months <- c(5:7) #focusing on May-Jul

snotel %>% filter(mm %in% months) %>%
  ggplot(aes(x = year, y = ppt_i_cm))+
  geom_point(aes(col = as.factor(mm))) #ok no big outliers. No missing values

snotel %>% filter(mm %in% months) %>%
  ggplot(aes(x = year, y = tavg_d_C))+
  geom_point(aes(col = as.factor(mm))) #removed 194 rows
#have two low values below -50, in Jun, Jul -> suspect incorrect since they're WAY below anything else seen (lowest is ~ -25 avg otherwise)

#removing low temp outliers (not month specific here since I only care about specific ones I've plotted)
snotel$tavg_d_C[which(snotel$tavg_d_C < -50)] <- NA

#group by year then month -> will aggregate across days and sites
sno_avg <- snotel %>% 
  select(-c(dd, site)) %>%
  group_by(year, mm) %>%
  summarize(across(.cols = everything(),
                   ~mean(.x, na.rm = TRUE),
                   .names = "avg_{.col}"))

#create mjj slices for each year
sno_mjj <- sno_avg %>% 
  #mjj only
  filter(mm >4 & mm < 8) %>%
  select(-mm) %>%
  group_by(year) %>%
  summarise(across(.cols = everything(),
                   ~mean(.x, na.rm = TRUE),
                   .names = "{.col}_mjj"))
```

Compare to PRISM climate data
```{r}
climAnnPrism <- read_csv("climAnnPrism.csv")

#compare date ranges
names(climAnnPrism)
summary(climAnnPrism$year)

#SNOTEL is more limited - 1986-2019 (ie lack older data)
#so, if it's fairly similar to PRISM, would be MORE IDEAL TO USE PRISM

#subsetting SNOTEL to just the key temperature variable & relabelling for clarity
sno_mjj <- sno_mjj %>% 
  select(year, "avg_mjj_tmean_C_snotel" = avg_tavg_d_C_mjj)

#joining SNOTEL & PRISM datasets
climAnnCompar <- sno_mjj %>%
  dplyr::inner_join(climAnnPrism, by = "year")

#compare temp estimates between climate sources
cor.test(climAnnCompar$avg_mjj_tmean_C_snotel, 
    climAnnCompar$avg_mjj_tmean_C) #0.97 correl - highly correlated

#plot correlation
(corr_plot <- climAnnCompar %>% ggplot(aes(x = avg_mjj_tmean_C, y = avg_mjj_tmean_C_snotel))+
  geom_point(col ="grey66")+
  geom_smooth(method = "lm", col = "black")+
  labs(x = "PRISM mean temperature (C)",
       y = "SNOTEL mean temperature (C)"))


#plot line graph of temp over time for both sources, to better see if they show similar trends

#melt data so easier to make a legend
(clim_raw_compar_plot <- climAnnCompar %>% 
  select(year, avg_mjj_tmean_C_snotel, avg_mjj_tmean_C) %>%
  pivot_longer(cols = c(avg_mjj_tmean_C_snotel, avg_mjj_tmean_C),
               names_to = "Source", values_to = "avg_mjj_tmean") %>% 
  ggplot(aes(x = year, y = avg_mjj_tmean, col = Source)) +
  geom_point()+
  geom_line()+
  labs(y = "Mean temperature\n(C, May - Jul)",
       x = "Year")+
  scale_color_manual(labels = c("PRISM", "SNOTEL"), 
                     values = c("black", "gray66")))
```

The temperature estimates from the two data sources are very highly correlated and show similar trends through time. SNOTEL temperature is consistently slightly lower than PRISM, except for one year ~ 2003.
  
Create a panel figure showing these two comparisons
```{r}
#export patchwork plot
pdf("prism snotel temp comparison - patchwork.pdf",
    width = 8, height = 4)

clim_raw_compar_plot + corr_plot

dev.off()
```